---
title: "RNAseq - SAMHD1 mouseGL261 Round2 - DESeq2 preprocessing"
author: Maike
date: '`r format(Sys.Date(), "%B %d, %Y")`'
---

**conda environment: conda activate R-4.2**


Set rmd options:
```{r setup1}
library(knitr)
library(rmarkdown)
knitr::opts_chunk$set(
  #collapse == TRUE,
  fig.width = 8,
  fig.height = 6
)
```



Load packages:
```{r load_packages, warning=FALSE}
# problem with tidyverse, now fine
#install.packages("tidyverse")

# load packages
suppressPackageStartupMessages({
  library(DESeq2)
  library(stringr)
  library(apeglm)
  # pca:
  library (limma)
  library(ggplot2)
  library(RColorBrewer)
  library(viridis)
  # circles:
  library(tidyverse)
  library(ggforce)
  # etc
  library(scales)
  library(dplyr)
})
```


# Data import

*Define output directory & load data:

*Import cts: from featureCounts output!

*Import metadata: manual input csv

This part will give a strange error. Ignore, just repeat this block an carry on.


```{r load_data}
# '/data2/240409_RKoenig_mouseGL261_round2/240411_featureCount/mouseGL261_readCounts_norm.csv'"), sep="\t", dec = ".", header=TRUE, row.names = 'Geneid')
ctsTot <- read.table(pipe("cut --complement -f2 '/home/AD/hermi/sm_tutorial/BYOC/testrun/240526_featureCount/mouseGL261_readCounts_norm.csv'"), sep="\t", dec = ".", header=TRUE, row.names = 'Geneid')

#/data2/240409_RKoenig_mouseGL261_round2/240409_RNAseq_mice_samples.csv'
metaFile <- '/home/AD/hermi/sm_tutorial/BYOC/myData/240409_RNAseq_mice_samples.csv'
coldataTot <- read.table(metaFile, sep="\t", dec = ".", header=TRUE)
```


# Prep count table

*remove spaces and '-'

*keep raw counts only

```{r prep count table}
# sample names: colnames(ctsTot): currently MedGen.LIB.ID
colnames(ctsTot) <- str_replace(colnames(ctsTot), "-", "_")
colnames(ctsTot) <- str_replace(colnames(ctsTot), " ", "_")
# feature names: rownames(ctsTot): leave alone.
#rownames(ctsTot) <- str_replace(rownames(ctsTot), "-", "_")
#rownames(ctsTot) <- str_replace(rownames(ctsTot), " ", "_")

# keep raw values only:
#   get lists of colnames without matching queries:
#   'negate=TRUE' selects non-matching strings only!
keep <- str_subset(colnames(ctsTot), pattern='TPM_', negate=TRUE)
keep <- str_subset(keep, pattern='RPKM_', negate=TRUE)
print(keep)
# filter:
ctsTot <- ctsTot[, keep]
identical(keep, colnames(ctsTot))
```


# Prep metadata table:

*remove spaces and '-'

*make coldataTotID match colnames(ctsTot):

```{r prep coldata}
# remove spaces and '-':
coldataTot <- coldataTot %>%
  mutate_all(str_replace_all, "-", "_") %>%
  mutate_all(str_replace_all, " ", "_")
print(colnames(coldataTot))

# make ID match ctsTot:
coldataTot$MedGen_ID_ed <- paste0('X', coldataTot$MedGen.LIB.ID)
#coldataTot$MedGen_ID_ed
#colnames(ctsTot)
coldataTot$MedGen_ID_ed <- str_replace(coldataTot$MedGen_ID_ed, "_SL1", "")
coldataTot$MedGen_ID_ed
#print(coldataTot$MedGen_ID_ed)
```


# Check correspondence cts & coldata:

```{r check correspondence}
# check:
identical(coldataTot$MedGen_ID_ed, colnames(ctsTot))
#[1] FALSE
setdiff(coldataTot$MedGen_ID_ed, colnames(ctsTot))
#[1] "X0246_2_TP04_005_FC_MOD_1TR1"

# order both df by MedGenId
coldataTot <- coldataTot[order(coldataTot$MedGen_ID_ed), ]
ctsTot <- ctsTot[, order(colnames(ctsTot))]
#colnames(ctsTot)

# or any sample not in counts table
keep <- colnames(ctsTot)
rownames(coldataTot) <- coldataTot$MedGen_ID_ed
coldataTot <- coldataTot[keep, ]
# check:
identical(coldataTot$MedGen_ID_ed, colnames(ctsTot))
#[1] TRUE

# list samples kept:
colnames(coldataTot)
coldataTot$Customer.ID
# ok

# Set row/column names
# use SampleNames(with batch), not MedGenIds: uniqie identifiers
rownames(coldataTot) <- coldataTot$Customer.ID
colnames(ctsTot) <- coldataTot$Customer.ID

# add lables column:
coldataTot$sLabel <- str_replace(coldataTot$Customer.ID, 'GL261_', '')
coldataTot$sLabel <- str_replace(coldataTot$sLabel, '_Batch_1', '_Batch1')
coldataTot$sLabel <- str_replace(coldataTot$sLabel, '_Batch1', '')
coldataTot$sLabel
```




# Prep dds input:

*Factorize: Reference lvl first, but here anything is fine: lavels may be adjusted later.

```{r prep dds}
#colnames(coldataTot)

unique(coldataTot$group)
coldataTot$group <- factor(coldataTot$group, levels = c('WT', 'Off_target_control', 'SAMHD1_KO') )
unique(coldataTot$group)

```


# Get dds object:

*set design & get matrix for DESeq2:

  variable of interest last, variable to account for (remove) first!
 
*for PCA: design doesn't matter:

  plotted are transformed readcounts, DGE is not conidered for DGE, ie, L2FC, design does matter!!!
  
  design = ~ donor + strain: correct for donor, test for DEG between strains

```{r get dds}
dds <- DESeqDataSetFromMatrix(countData = ctsTot,
                              colData = coldataTot,
                              design = ~ group)
dds
```


# Save dds:

```{r save dds}
outFile <- '240526_rawDds.rds'
saveRDS(dds, file = outFile)
# restore:
#dds <- readRDS(outFile)
```


# Get session info:

```{r session info}
sessionInfo()
```







