---
title: "DEG & PCA - SAMHD1 mouseGL261 Round2 - Test1: off_target_control vs. SAMHD1_KO vs. WT"
author: Maike
date: '`r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  chunk_output_type: inline
---


**conda environment: conda activate R-4.2cpFix**


```{r setup1}
library(knitr)
library(rmarkdown)
knitr::opts_chunk$set(
  #collapse == TRUE,
  fig.width = 8,
  fig.height = 6
)
```

Load packages:
```{r load_packages, warning=FALSE}
# problem with tidyverse, now fine
#install.packages("tidyverse")

# load packages
suppressPackageStartupMessages({
  library(DESeq2)
  library(stringr)
  library(apeglm)
  # pca:
  library (limma)
  library(ggplot2)
  library(RColorBrewer)
  library(viridis)
  # circles:
  library(tidyverse)
  library(ggforce)
  # etc
  library(scales)
  library(dplyr)
  library(xlsx)
})
```


# Define output directory & load data:
**input from 
/data2/240409_RKoenig_mouseGL261_round2/240411_DESeq2/240411_DESeq2_preprocessing.rmd**

dds with raw read counts of all samples

```{r load_data}
# setup:
inFile <- '/data2/240409_RKoenig_mouseGL261_round2/240411_DESeq2/240411_rawDds.rds'
dds <- readRDS(inFile)
dim(dds)
```


## TEST1: Off_target_control (n=3) vs. SAMHD1_KO (n=3) vs. WT (n=1)

# Prep dds input: Factorize: done in preprocessing

Reference lvl first! = WT!

*order group


# update dds & deseq2:

* not necessary


# Apply read count filter

theoretically unneccessary

* set prefilter for included genes to min no of reads (depends...)

* in min no of samples (usually number of samples in smallest group)

* here: min 1 sample with at least 5 reads (there are groups with n=1)

```{r readcount filter dds}
smallestGroupSize <- 1
keep <- rowSums(counts(dds) >= 5) >= smallestGroupSize 
# apply filter:
dds <- dds[keep, ]
dim(dds)
```


# PCA

Run DESeq2 for PCA: simple design is fine, doesn't matter

PCA is based on raw counts!

* DESeq2 on dds

* vst on dds
```{r DESeq2 vst}
dds <- DESeq(dds)
#get Ids:
resultsNames(dds)

# transform for PCA
vsd <- vst(dds, blind=FALSE)
vsd
# check vsd data:
#head(assay(vsd))
```


# plot PCA

* get raw PCA

```{r PCA}
colnames(dds@colData)
# include metadata of interest:
pca <- plotPCA(vsd, intgroup=c("sLabel", 
                               "group"), returnData = FALSE)
print(pca)
```

* get PCA metadata

```{r PCA metadata}
# pca data frame:
#pca$data
# pca labels
pca$labels
xTitle <- pca$labels[[2]]
yTitle <- pca$labels[[1]]
# col:
myFills <- scales::viridis_pal()(3)
#show_col(myFills)
# save dimensions as text:
geneNo <- as.character(dim(dds)[1])

```


* ggplot DESeq2 plotPCA plot

```{r ggplot PCA}
pTitle <- "PCA GL261 SAMHD1 - Test 1: Off target control vs. KO vs WT"
pFoot <- paste0(c("\nno batch effect present",
                  "\nprefiltering applied:  min 5 reads/gene in min 1 sample",
                  "\nretained features: ", geneNo), collapse='')
pFoot
pcaPlot <- 
  ggplot(data = pca$data, aes(x=PC1, y=PC2, group = group, label=sLabel)) + 
  #geom_point(aes(color = batch, shape = mutation), size = 2, stroke=2) +
  geom_point(aes(color = group.1, shape = group.1), size = 3, stroke=2) +
  # nudge moves label; 
  # h/vjust are for aligning text in box: left/right/center/top/bottom!
  #geom_text(angle=45) + # don't nudge prefiltered data! it's too much
  geom_text(nudge_y = 0.6, nudge_x = 0, hjust = 0.2, vjust = 0, size = 3, color = 'grey25') + #, check_overlap = TRUE removes labels!
    #geom_label(nudge_y = 1, nudge_x = 15)
  scale_color_manual(values = myFills) +
  # preserve aspect ratio: 1 unit on x is as long as 1 unit on y
  coord_fixed() +
  #lims(y=c(-0.2, 0.2)) + 
  labs(title = pTitle,
       #subtitle = subTitle,
       caption = pFoot,
       x = xTitle, y = yTitle) +
  theme_bw() +
  theme(
    # remove grid:
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    # remove background:
    panel.background = element_blank(),
    # control legend spacing:
    legend.key.size = unit(0.8, 'cm'))
print(pcaPlot)
```

# Save PCA to pdf
```{r save PCA}

#getwd()
pcaOut <- '240411_Test1_PCA_CtrlvsKOvsWT.pdf'
# the aspect ratio is preserved automatically! It might seem a bit of but that is an illusion!!!
cairo_pdf(filename = pcaOut,
          width = 12, 
          height = 6, 
          pointsize = 12)
print(pcaPlot)
dev.off()
```




# DE analysis tables

3-wise comparison vs WT

```{r contrast MAP function}
# get Ids:
resultsNames(dds)
length(resultsNames(dds))

get_contrastMAP_tab <- function(dds, idx){
  # specify contrast
  # get apeglm MAP values
  # edit result: data frame, columnnames
  IDs_dds <- resultsNames(dds)
  print(paste0('contrasts: ', paste(IDs_dds, collapse='; ')))
  id <- IDs_dds[idx]
  print(paste0('now computing ', id))
  # get MAP & keep df for output table:
  res <- lfcShrink(dds, coef=id)
  res <- as.data.frame(res)
  # edit colnames for output table: APPLY AFTER PLOTTING!!!
  colnames(res) <- paste(id, colnames(res), sep='_')
  return(res)
}
```


3-wise comparison vs Off_target_control

* relevel dds

* new test

```{r get last contrast}
# relevel
dds$group <- relevel(dds$group, 'Off_target_control')
unique(dds$group)
dds2 <- nbinomWaldTest(dds)
# takes a while, but less than from scratch.
resultsNames(dds2)
```

# get MAP values

*The last contrasts results are not based on the same distribution as the other two. Shrinking is affected by the reference sample choice!*

The difference is, however, small.

```{r get contrasts}
con5 <- get_contrastMAP_tab(dds, 2)
con6 <- get_contrastMAP_tab(dds, 3)
con7 <- get_contrastMAP_tab(dds2, 3)
```

# save MAP tables

```{r save MAP values}

prefix='240411_Test1_DESeq2_'

idx <- 2
con <- con5
ddstab <- dds
baseName <- paste0(c(prefix, resultsNames(ddstab)[idx]), collapse = '' )
baseName <- str_replace(baseName, '_group', '')
baseName
#   col.names=NA, row.names=TRUE) avoids columnnames starting in first field (above genenames):
write.table(con, file=paste0(baseName, '.csv'), sep='\t', col.names=NA, row.names=TRUE)
# xlsx output
write.xlsx(con, file=paste0(baseName, '.xlsx'))

idx <- 3
con <- con6
ddstab <- dds
baseName <- paste0(c(prefix, resultsNames(ddstab)[idx]), collapse = '' )
baseName <- str_replace(baseName, '_group', '')
baseName
#   col.names=NA, row.names=TRUE) avoids columnnames starting in first field (above genenames):
write.table(con, file=paste0(baseName, '.csv'), sep='\t', col.names=NA, row.names=TRUE)
# xlsx output
write.xlsx(con, file=paste0(baseName, '.xlsx'))

idx <- 3
con <- con7
ddstab <- dds2
baseName <- paste0(c(prefix, resultsNames(ddstab)[idx]), collapse = '' )
baseName <- str_replace(baseName, '_group', '')
baseName
#   col.names=NA, row.names=TRUE) avoids columnnames starting in first field (above genenames):
write.table(con, file=paste0(baseName, '.csv'), sep='\t', col.names=NA, row.names=TRUE)
# xlsx output
write.xlsx(con, file=paste0(baseName, '.xlsx'))
```

get session info:
```{r session info}
sessionInfo()
```


