---
title: "Volcano plots - SAMHD1 mouseGL261 Round2 - Test1.1: off_target_control vs. SAMHD1_KO vs. WT without sample GL261_AF_2_3_Batch1"
author: Maike
date: '`r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  chunk_output_type: inline
---

**conda environment: conda activate R-4.2**


```{r setup1}
library(knitr)
library(rmarkdown)
knitr::opts_chunk$set(
  #collapse == TRUE,
  fig.width = 8,
  fig.height = 6
)
```

Load packages:
```{r load_packages, warning=FALSE}
# problem with tidyverse, now fine
#install.packages("tidyverse")

# load packages
suppressPackageStartupMessages({
library(dplyr)
library(pathview)
library( "stringr" )
library( "ggplot2" )
library( "RColorBrewer" )
library( "EnhancedVolcano" )
library(data.table)
library(gridExtra)
library(grid)
library(cowplot)
})
```


# functions

* added hallmark genes (Rice et al. 2013)

```{r functions}
# functions:
# -------------
# -------------
get_rawDF <- function(resFile){
  # BEWARE OF THE COLUMNNAMES!!!!
  # input file = DESeq2 results
  # !!! import DGE results table, 
  # !!! filter, order:  keep l2FC, padj, pval, gene name
  res <- read.table(resFile, sep="\t", dec = ".", header=TRUE, na.strings = 'NA')
  res$SYMBOL <- res$X
  
  # make gene names row names unless this breaks something
  rownames(res) <- res$SYMBOL
  
  lfcCols <- str_subset(colnames(res), pattern='log2FoldChange')
  paCols <- str_subset(colnames(res), pattern='padj')
  pCols <- str_subset(colnames(res), pattern='pvalue')
  #print(pCols)
  keepLFC <- c(lfcCols, paCols, pCols, 'SYMBOL')
  # df <- df[rows, columns]
  res_ed <- res[, keepLFC]
  return(res_ed)
}

get_conDf <- function(res_ed, contrast){
  # BEWARE OF THE COLUMNNAMES!!!! 
  #   order is set in get_rawDF: l2FC, padj, pval, gene name
  # automatically pick contrast & default columns (l2FC , padj) then rename columns
  keepCols <- c('SYMBOL', contrast)
  keepPat <- paste(keepCols, collapse = '|')
  keep <- grep(keepPat, colnames(res_ed))
  conDf <- res_ed[, keep]
  # rename columns
  colNames <- c('log2FoldChange',
                 'padj',
                 'del1',
                 'del2')
  colnames(conDf) <- colNames
  # for volcanoplot: filter and rename again
  #   not very slick, but whatever
  keep <- c("log2FoldChange", "padj")
  conDf <- conDf[, keep]
  return(conDf)
}

make_GeneCounts <- function(VolPlot, colList) {
  # get gene counts each category
  # --------------------
  # '-log10 p' is roughly providing the exponent-1 used in scientific notation 
  #   e.g. p = 0.000045 = 4,5*(10â»5); -log10 p = 4.3 (= close to 4)
  # ------------------------------------------------------
  build <- ggplot_build(VolPlot)
  build_data <- as.data.frame(build$data[1])
  # 1. ns but l2fc low
  CountNsLo <- nrow(subset(build_data, colour == colList[1]))
  # 2. ns & l2fc high
  CountNsHi <- nrow(subset(build_data, colour == colList[2]))
  # 3. sig & l2fc low
  CountSiLo <- nrow(subset(build_data, colour == colList[3]))
  # 4. sig & l2fc high
  CountSiHi <- nrow(subset(build_data, colour == colList[4]))
  gCounts <- c(CountNsLo, CountNsHi, CountSiLo, CountSiHi)
  print(paste('plot-based total gene counts:', sum(gCounts)))
  return(gCounts)  
}

make_volPlot <- function(res, L2FCTH, yMax, legLabs) {
  # plot area, legend = gene counts
  volPlot <- EnhancedVolcano(res,
                             lab = rownames(res),
                             x = 'log2FoldChange',
                             y = 'padj',
                             # remove all titles
                             title = '',
                             subtitle = '',
                             caption = '',
                             legendLabels = legLabs,
                             legendDropLevels = FALSE,
                             ylab = bquote(~-Log[10]~italic(padjust)),
                             ylim = c(0, yMax),
                             xlim = c(-10, 10),
                             pCutoff = 0.05,
                             FCcutoff = L2FCTH,
                             pointSize = 3.0,
                             
                             # GENE SYMBOLS HERE!!! 
                             ## IMPORTANT: case sensitive!!!
                             #selectLab = '',
                             selectLab = c('Usp18', 'Ifi44', 'Oasl1', 'Rsad2', 'Samhd1'),
                             #selectLab = c('IFI27', 'IFI44L', 'ISG15', 'IFIT1', 'RSAD2', 'SIGLEC1'),
                             drawConnectors = TRUE,
                             labSize = 6.0,
                             gridlines.major = FALSE,
                             gridlines.minor = FALSE,
                             col =  c('grey',"#66C2A5", "#8DA0CB", "#E78AC3", "#A6D854", "#FC8D62" )) +
    #selectLab = c('')) +
    # set legend title
    guides(color = guide_legend(title = paste(nrow(res), 'genes'))) +
    theme(# make area square 
      aspect.ratio = 1,
      # adjust legend: location, title, outline
      legend.position = c(0.2, 0.8),
      legend.title = element_text(),
      legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid', linewidth = 0.5),
      # control plot margins:
      plot.title = element_blank(),
      plot.subtitle = element_blank(),
      plot.caption = element_blank(),
      plot.margin = margin(0,0,0,0, 'cm'))
  # remove legend
  #legend.position = 'none')
  print(volPlot)
} 

make_legPlot <- function(res, L2FCTH, legLabs) {
  # plot for descriptive legend:
  volPlot <- EnhancedVolcano(res,
                             lab = rownames(res),
                             x = 'log2FoldChange',
                             y = 'padj',
                             ylim = c(0, 100),
                             xlim = c(-10, 10),
                             # remove some titles
                             title = '',
                             subtitle = '',
                             #caption = '',                             
                             # high cutoff - to get all legend items
                             pCutoff = 0.10,
                             FCcutoff = L2FCTH,
                             pointSize = 3.0,
                             labSize = 6.0,
                             gridlines.major = FALSE,
                             gridlines.minor = FALSE,
                             legendLabels = legLabs,
                             col =  c('grey',"#66C2A5", "#8DA0CB", "#E78AC3", "#A6D854", "#FC8D62" ),
                             #legendPosition = 'bottom')
                             legendPosition = 'right')
  #theme(legend.background = element_rect(colour = 'black', fill = 'grey', linetype='solid', linewidth = 0.5))
  print(volPlot)
}

```


# plot settings: colors etc.

```{r volcano settings}
dummyLabs <- c(1, 2, 3, 4)
myColors <- c('grey',"#66C2A5", "#8DA0CB", "#E78AC3", "#A6D854", "#FC8D62" )
legLabs=c('n.s. & low L2FC', 'n.s. & L2FC > 1', 'p <= 0.05 & low L2FC', 'p <= 0.05 & L2FC > 1')
```


# set contrasts and input files

* input files were generated with
/data2/240409_RKoenig_mouseGL261_round2/240411_DESeq2/240411_test1.1_DESeq2_WTvsKOvsCtrl_noAF2.3.rmd*

* "cons"" must be the actual contrast Ids used in the csv header fields!!! Juste take it from file name.

```{r input settings}
conLabs <-  c('Off target control vs. WT',
              'SAMHD1 KO vs. WT',
              'SAMHD1 KO vs. Off target control')

cons <-  c('Off_target_control_vs_WT',
           'SAMHD1_KO_vs_WT',
           'SAMHD1_KO_vs_Off_target_control')

#conFiles <- c('/data2/240409_RKoenig_mouseGL261_round2/240411_DESeq2/240411_Test1.1_DESeq2_Off_target_control_vs_WT.csv',
#              '/data2/240409_RKoenig_mouseGL261_round2/240411_DESeq2/240411_Test1.1_DESeq2_SAMHD1_KO_vs_WT.csv',
#              '/data2/240409_RKoenig_mouseGL261_round2/240411_DESeq2/240411_Test1.1_DESeq2_SAMHD1_KO_vs_Off_target_control.csv')
conFiles <- c('/home/AD/hermi/sm_tutorial/BYOC/testrun/DESeq2/240526_Test1.1_DESeq2_Off_target_control_vs_WT.csv',
              '/home/AD/hermi/sm_tutorial/BYOC/testrun/DESeq2/240526_Test1.1_DESeq2_SAMHD1_KO_vs_WT.csv',
              '/home/AD/hermi/sm_tutorial/BYOC/testrun/DESeq2/240526_Test1.1_DESeq2_SAMHD1_KO_vs_Off_target_control.csv')

```


# plot:

EnhancedVolcano tries to label all genes passing the tresholds

here: pCutoff = 0.05, FCcutoff = 1

It labes only as many genes as to be still readable

Order of Graphs is messed up in Rstudio preview but fine in html output

Check yMax!!!

```{r volcano loop}
dummyList <-list()
maxY <- 100

yMaxList <- vector()
plotList <- list()

# all:
for(i in 1:length(cons)) {
  # get raw df:
  inFile <- conFiles[i]
  res_tab <- get_rawDF(inFile)
  # gene names: res_tab$SYMBOL
  
  # set contrast and get DF contrast, L2FC + padj only:
  con <- cons[i]
  print(paste('plotting', con))
  vol_df <- get_conDf(res_tab, con)
  
  # pval - yAxis range:
  vol_df$negl10 <- -(log10(vol_df$padj))
  yRange <- c(min(vol_df$negl10, na.rm = TRUE), max(vol_df$negl10, na.rm = TRUE))
  ymax <- yRange[2]
  print(paste(con, 'yMax:', ymax))
  
  # get gene counts & percentages per category
  dummyPlot <- make_volPlot(vol_df, 1, maxY, dummyLabs)
  gCountsRaw <- make_GeneCounts(dummyPlot, myColors)
  gCountsPerc <- paste(round((gCountsRaw / nrow(vol_df) * 100), 1), '%')
  gCounts <- paste0(gCountsRaw, ' (', gCountsPerc, ')')
  # plot; legend = gene counts
  vPlot_p1 <- make_volPlot(vol_df, 1, maxY, gCountsPerc)
  # adjust x axis limits:
  #print(paste0(conLabs[i], '\n'))
  vPlot_p1 <- vPlot_p1 + xlim(-10, 10) +
    labs(title = paste0(conLabs[i], '\n')) +
    theme(plot.title = element_text(),
          # legend to the right:
          legend.position=c(0.9, 0.8),
          plot.margin = margin(0.2,0,0,0, 'cm'))
  print(vPlot_p1)
  # check whether legend covers dots:
  print(vPlot_p1 + theme(legend.position='none'))
  # append to lists:
  yMaxList <- c(yMaxList, ymax)
  plotList[[i]] <- vPlot_p1
}
```



```{r check loop results}
# check whether all dots are included (bc yaxis limit)
# all:
#for(i in 1:length(cons)) {
#  con <- cons[i]
#  print(con)
#  vPlot_p1 <- 0
#  vPlot_p1 <- plotList[[i]]
#  # print wo legend:
#  print(vPlot_p1 + theme(legend.position='none'))
#  # print with legend:
#  print(vPlot_p1)  
#}
```




```{r legend loop}
# --------------------------------
# plot; legend = categories (only once!!)
# --------------------------------
leg <- make_legPlot(vol_df, 1, legLabs) +
  theme(legend.text = element_text(size = 10),
        legend.position = c(0.5, 1.2))
  #guides(fill=guide_legend(nrow=2, byrow=TRUE)) #+
  #theme(legend.margin = unit(-3, 'cm'))
leg <- get_legend(leg)
# legend to grid:
leg <- plot_grid(NULL, leg, ncol=1)
# legend to grob:
leg <- as_grob(leg)
# display leg:
grid.draw(leg)

```


```{r save plot}
# test:
id=1

foot <- "no batch effect present; sample AF_2_3 removed"
grid.arrange(plotList[[i]], leg , nrow = 1, widths = c(10, 4),
             bottom = textGrob(foot, 
                               gp = gpar(fontsize = 9),
                               hjust = 1,
                               y=1,
                               x = 0.9))

# loop:
for(i in 1:length(cons)) {
  plotOut <- paste0('240411_Test1.1_volcano_l2FCTH1_', cons[i], '.pdf')
  print(plotOut)
  cairo_pdf(filename = plotOut,
            width = 8, 
            height = 7, 
            pointsize = 12)
  grid.arrange(plotList[[i]], leg , nrow = 1, widths = c(10, 4),
               bottom = textGrob(foot,
                  gp = gpar(fontsize = 9),
                               hjust = 1,
                               y = 1,
                               x = 0.9))
  dev.off()
}
```




get session info:
``` {r session info}
sessionInfo()
```


